---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r loadlibraries, echo=F}
library(data.table)
library(ggplot2)
library(xtable)
```

##Reproducible Research: Peer Assessment 1. Activity monitor analysis

## Loading and preprocessing the data
The data to work with is included in this github project, the first step would
be to unzip the data and place it under the `data/` directory  and load it into
R. **Note**: the `data` directory won't be commited to github.

```{r loaddata}
unzip('activity.zip', exdir='data/')
data <- fread('data/activity.csv')
data$date <- as.Date(data$date, format='%Y-%m-%d')
```

## What is mean total number of steps taken per day?
We add the total number of steps per day and group them by the date.
A histogram is created to illustrate the total number of steps.

```{r stepsbyday}
totalStepsByDay <- data[,.(total=sum(steps)), .(date)]
ggplot(totalStepsByDay, aes(x=date, weight=total)) +
    geom_histogram(binwidth=1) +
    labs(title='Total steps by day', y='Steps')
```

We will also report the mean and median steps taken by day. We first ignore the
entries with `NA` steps.

```{r meanstepsbyday}
stepsByDay <- data[!is.na(steps),
                   .(
                     meanDay=mean(steps, na.rm=T),
                     medianDay=median(steps, na.rm=T)
                     ), .(date)]
```

Notice that some dates are missing, this is because there's no data in those days.

```{r tablemeanstepsbyday, results='asis', echo=F}
stepsByDay$date <- as.character(stepsByDay$date)
tableStepsByDay <- xtable(stepsByDay)
print(tableStepsByDay, type='html', include.rownames=F)
```

## What is the average daily activity pattern?
We want to know on average how many steps are given on a time period. Remember,
the data set contains a variable called `interval` which is the 5 minute period
where the read was taken, for example, 0 represents 0:00, 1455 represents
14:55, and so on.

In this case we calcuated the average steps grouped by the interval across the
entire data set. The plot displays the mean steps for interval 1000; to calculate
this, all existing steps in the 1000 `interval` were averaged.

```{r intervalmean}
meanByInterval <- data[,.(avgSteps=mean(steps, na.rm=T)), .(interval)]
ggplot(meanByInterval, aes(interval, avgSteps)) +
    geom_line() +
    labs(title='Average steps by interval', y='Steps')
```

```{r maxinterval, echo=F}
maxStepsInterval <- meanByInterval[which.max(avgSteps), .(interval, avgSteps)]
interval <- maxStepsInterval$interval
avgSteps <- maxStepsInterval$avgSteps
```

We can see that the interval with the max average steps is the `r interval`
interval with an average of `r avgSteps`.

## Imputing missing values

```{r totalNas, echo=F}
totalNas <- data[is.na(steps), .N]
```

This data set contains a total of `r totalNas` `NA` values in the steps variables.
In order to fix this we will replace those `NA`s with the median of the day. In
the days where the median cannot be calculated (there are no reading for that
specific day) we will set the steps to 0.

Going from `NA`s to 0 is probably not the optimal. One improvement for this would be
to calculate the mean of the observations of other days with observation,
I'd assume that a similar routine is followed for example in every Monday.

The steps to replace `NA`s with the media is as follow.

1. Replicate our data set.
2. Detect the days that have `NA`s in it.
3. Apply the function `replaceNAs` which will do the following
   1. It will get the median using the mean/median set displayed above.
   2. If the row doesn't exist we default to a median 0
   3. The selected median is applied to all `NA`s steps.

```{r replacingnas}
replaceNAs <- function(dayDate) {
    dayMedian <- 0
    steps <- stepsByDay[date==dayDate]

    # If we don't have a row, then  day median is NA (there are no steps recorded
    # in the whole day, we set the rowMedian to 0
    if (nrow(steps) > 0) {
        dayMedian <- steps$medianDay
    }

    imputed[date==dayDate & is.na(steps), steps:=dayMedian]
}

# We will create a copy of the data
imputed <- data.table(data)

# We detect the dates with NAs in it
nasDates <- imputed[is.na(steps), unique(date)]

a <- lapply(nasDates, replaceNAs)
```

To see the histogram of this values we add the steps and display.

```{r plotimputed}
imputedTotals <- imputed[,.(total=sum(steps)), .(date)]
ggplot(imputedTotals, aes(x=date, weight=total)) +
    geom_histogram(binwidth=1) +
    labs(title='Total steps by day', y='Steps')

```

What's the difference between this histogram and the previous without the `NA`s
replace? We will take both histograms data and calculate the difference. The
table below shows the results.

```{r stepsdifference}
totalsDifference <- merge(totalStepsByDay, imputedTotals, by='date')
fixedDiff <- totalsDifference[,
              .(date,
                diff=abs(ifelse(is.na(total.x), 0, total.x) - total.y)
                )]
```

Basically, replacing `NA`s with the median, wouldn't change the results that
we've seen so far.

```{r tablediff, results='asis', echo=F}
fixedDiff$date <- as.character(fixedDiff$date)
tableDiff <- xtable(fixedDiff)
print(tableDiff, type='html', include.rownames=F)
```

## Are there differences in activity patterns between weekdays and weekends?
First we will classify weekday and weekends, `as.POSIXlt(date)$wday` will
return a number 1=Monday, 2=Tuesday and so on. Any day bigger than 5=Friday
will be considered weekend.

```{r weekdays}
imputed[,weekday:=as.factor(ifelse(as.POSIXlt(date)$wday > 5, 'weekend', 'weekday'))]
```

Once we have the days classified as weekend/weekday we will add the total steps
by interval.

```{r weekdaysplot}
weekDay <- imputed[,.(total=mean(steps)), .(interval, weekday)]
ggplot(weekDay, aes(interval, total)) +
    geom_line() + 
    facet_grid(weekday ~ .) +
    labs(title='Total steps by interval on day type', y='Total steps')
```
